noam_data_exploration_v1

### 2024-07-18 noam_data_exploration_v3.Rmd
Subsetted quad level data according to what Lauren did (I messed it up somehow) and ran total turnover + some exploratory stats at quad level.

problem_children <- c("LTER 5 Fringing Reef Algae Transect 2 Quad 9", "LTER 5 Fringing Reef Algae Transect 3 Quad 9", "LTER 5 Fringing Reef Algae Transect 3 Quad 8", "LTER 5 Fringing Reef Algae Transect 3 Quad 6", "LTER 5 Fringing Reef Algae Transect 3 Quad 3", "LTER 5 Fringing Reef Algae Transect 3 Quad 10")

long_data_sub <- long_data[long_data$Location != problem_children, ] # remove problem children
# still get the same error when I try to run at quadrat level
# I think it's because of the 0s
# Error in names(output) <- `*vtmp*` : 
#  'names' attribute [1] must be the same length as the vector [0]
```

```{r turnover, eval = FALSE}
# the turnover function calculates turnover as the proportion or species that either appear or disappear between timepoints
total_turnover<-
  turnover(
  df = long_data_sub,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Percent_cover",
  replicate.var = "Location",
  metric = "total"
)

total_turnover<- 
  total_turnover %>%
    separate(Stratum, into = c("Site", "Habitat", "Transect"), sep = "-")
total_turnover$Site_habitat <- paste(total_turnover$Site, total_turnover$Habitat)
# Add back in stratum column

total_turnover <- total_turnover %>%
  mutate(across(everything(), str_trim))

str(total_turnover) # everything became characters

total_turnover$total <- as.numeric(total_turnover$total)
total_turnover$Year <- as.factor(total_turnover$Year)
total_turnover$Site <- as.factor(total_turnover$Site)
total_turnover$Habitat <- as.factor(total_turnover$Habitat)

turnover_means <-
  total_turnover %>%
  dplyr::group_by(Year, Site, Habitat, Site_habitat) %>%
  dplyr::summarise(Turnover_mean = mean(total), 
            Turnover_se = sd(total)/sqrt(n()))

(turnover_plot <-
  ggplot(data = turnover_means) +
  geom_point(aes(x = Year, y = Turnover_mean, colour = Site, shape = Habitat, group = Site_habitat), size = 8, stroke = 2) +
  geom_line(aes(x = Year, y = Turnover_mean, colour = Site, group = Site_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Turnover_mean - Turnover_se , ymax = Turnover_mean + Turnover_se, colour = Site,
                    group = Site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site_habitat, nrow = 6)+
  labs(title = "a. Turnover\n", y = "Total turnover\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )
#

# examine appearance and disappearances independently 
# refers to the proportion of species that appeared or disappeared relative to total species pool between two time points

>>>>

problem_children <- c(
                      "LTER 5 Fringing Reef Algae Transect 3 Quad 9",
                      "LTER 5 Fringing Reef Algae Transect 3 Quad 8",
                      "LTER 5 Fringing Reef Algae Transect 3 Quad 6",
                      "LTER 5 Fringing Reef Algae Transect 3 Quad 3",
                      "LTER 5 Fringing Reef Algae Transect 3 Quad 10")

long_data_sub <- long_data[!long_data$Location %in% problem_children, ] # remove problem children

data_sub <- data[!data$Location %in% problem_children, ]

```{r total turnover, fig.height = 24, fig.width=16}
# the turnover function calculates turnover as the proportion or species that either appear or disappear between timepoints
total_turnover<-
  turnover(
  df = long_data_sub,
  time.var = "Year",
  species.var = "Taxa",
  abundance.var = "Percent_cover",
  replicate.var = "Location",
  metric = "total"
)

# get relevant metadata for turnover metrics
data_sub <- data[paste(data$Year, data$Location) %in% paste(total_turnover$Year, total_turnover$Location), ]

# check to make sure everything is the same length
# nrow(data_sub) # 17829
# nrow(total_turnover) # 17829

# extract metadata
turnover_meta <- data_sub[ ,which(names(data_sub) %in% meta_cols)]

# merge with turnover df
total_turnover <- merge(total_turnover, turnover_meta, by = c("Year", "Location"))
total_turnover$Year_factor <- as.factor(total_turnover$Year)

# turnover bounded between 0 and 1 --> beta regression
# need to transform due to presence of 0s and 1s
# Smithson verkuilen transformation for beta-regression
sv_trans = function(prop, N, s = 0.000005){
  (prop*(nrow(N) - 1) + s)/nrow(N)
  # where prop is the proportional value you're transforming, 
  # N is the sample size, which is specified by taking the number of rows/observations from a given dataframe,
  # and s is a small offset 
}
total_turnover$Turnover_trans <- sv_trans(total_turnover$total, total_turnover)

# model
total_turnover_mod1 <- glmmTMB(Turnover_trans ~ Year_factor*Habitat*Site +
                                 (1 | Site/Transect/Location) + ar1(Year_factor + 0 | Location),
                               family = beta_family(), data = total_turnover)
summary(total_turnover_mod1) # low temporal autocorrelation... maybe don't need ar1 term?
plot(total_turnover_res <- simulateResiduals(total_turnover_mod1)) # residuals look very weird
Anova(total_turnover_mod1) # obviously everything sig

total_turnover_mod2 <- glmmTMB(Turnover_trans ~ Year_factor*Habitat*Site + (1 | Site/Transect/Location),
                               family = beta_family(), data = total_turnover)
summary(total_turnover_mod2) # 
plot(total_turnover_res2 <- simulateResiduals(total_turnover_mod2)) # residuals still look like shit
Anova(total_turnover_mod2) 
anova(total_turnover_mod2, total_turnover_mod1) # models are sig different from each other
AICc(total_turnover_mod2, total_turnover_mod1) # first model is better
# need to figure out residuals problem
plot(allEffects(total_turnover_mod1))

# plot
total_turnover_means <-
  total_turnover %>%
  dplyr::group_by(Year, Island_side, Site, Habitat, Site_habitat) %>%
  dplyr::summarise(Turnover_mean = mean(total), 
            Turnover_se = sd(total)/sqrt(n()))

(turnover_plot <-
  ggplot(data = total_turnover_means) +
  geom_point(aes(x = Year, y = Turnover_mean, colour = Site, shape = Habitat, group = Site_habitat), size = 8, stroke = 2) +
  geom_line(aes(x = Year, y = Turnover_mean, colour = Site, group = Site_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Turnover_mean - Turnover_se , ymax = Turnover_mean + Turnover_se, colour = Site,
                    group = Site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site_habitat, nrow = 6)+
  labs(title = "a. Turnover\n", y = "Total turnover\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )
#
```

```{r appearances}
# examine appearance and disappearances independently 
# refers to the proportion of species that appeared or disappeared relative to total species pool between two time points
